---
title: "Chantarat SR Measure"
author: "Tracy Lam-Hine"
date: "2023-01-13"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown

### Overview of Multidimensional Measure of Structural Racism (MMSR)

[Chantarat et al. 2021](https://doi.org/10.1016/j.eclinm.2021.101092) describe the details of their multidimensional measure of structural racism (MMSR) in the supplementary material file included in the [paper appendix](https://ars.els-cdn.com/content/image/1-s2.0-S2589537021003722-mmc1.pdf). In particular, they write:

> *We measured structural racism in five key dimensions at the Public Use Microdata Area (PUMA) (n=2,338 out of 2,351 PUMAs in the United States). All unidimensional structural racism measures below were derived using the 2019 American Community Survey (ACS) five-year estimates. "White" refers to non-Hispanic Whites, while "Black" refers to Blacks/African Americans, including both Hispanic and non-Hispanic. We were unable to use non-Hispanic Blacks/African Americans alone to derive our unidimensional measures because the ACS does not publish data for that group.*
>
> -   ***Black-White segregation:** measured by the index of dissimilarity (D). This index can be interpreted as the percentage of a group's population that would have to change residence to achieve a uniform distribution of the population. This index ranges from 0 (complete lack of segregation) to 1 (complete segregation).*$$ D = \frac{1}{2} \sum^{n}_{i=1} (\frac{w_i}{W_T} - \frac{b_i}{B_T})$$
>
>     *where* $n$ *= number of census tracts;* $w_i$ *= number of Whites in tract* $i$*;* $W_T$ *= total number of Whites in the PUMA;* $b_i$ *= number of Blacks in tract* $i$*;* $B_T$ *= total number of Blacks in the PUMA.*
>
> -   ***Education inequity:** measured by the ratio of White to Black college education (i.e., Bachelor's degree or higher) rates among individuals aged 25 years and over.*
>
> -   ***Employment inequity:** measured by the ratio of White to Black employment rates among civilians in the labor force aged 25-64 years.*
>
> -   ***Income inequity:** measured by the index of concentration at the extremes (ICE):* $ICE = \frac{A-P}{T}$, where $A$ *= number of White households with income of \$100,000 or higher (privileged group);* $P$ *= number of Black households with income lower than \$25,000 (deprived group);* $T$ *= total number of Black and White households in the PUMA. This index ranges from -1 (all households are in the deprived group) to 1 (all households are in the privileged group).*
>
> -   ***Homeownership inequity**: measured by the ratio of White to Black homeownership rates.*
>
> *All unidimensional measures were dichotomized into high and low structural racism based on their sample median.*

### Using `tidycensus` to get the data

The [`tidycensus`](https://walker-data.com/tidycensus/index.html) package provides a useful interface to Census and American Community Survey (ACS) data sets; we will use this package to load the appropriate data to construct the MMSR.

We will use the `get_acs` function to access aggregated 5-year ACS tables (as opposed to individual-level data) to calculate the unidimensional measures in the MMSR. For the segregation measure, we need counts at both the census tract and public use microdata area (PUMA) level. Tracts nest completely within a PUMA. Unfortunately, because the 2020 and 2021 5-year ACS releases use 2020 tract and 2010 PUMA boundaries, we need to use the 2015-2019 5-year ACS release to make sure our tracts and PUMAs align. When 2020 PUMAs are released sometime in 2023, we can recalculate the MMSR with the 2018-2022 5-year ACS data.

```{r load-packages, message = F}
library(tidyverse)
library(tidycensus)
library(data.table)
library(tigris)
library(mapview)
library(poLCA)
```

To access Census and ACS data using `tidycensus` you will need to obtain and set a Census API key in your R environment. You can obtain a key [here](https://api.census.gov/data/key_signup.html) - make sure not to share it with anyone.

```{r specify-key, eval = F}
# note - you will have to get a census API key in order to access census data 
census_api_key("YOUR API KEY GOES HERE")
```

```{r personal-key, include = F}
census_api_key("f77de7a3ed4802ce5f3a28b99ba1105b8bb39529")
```

#### Find our variables in the codebook for ACS aggregate tables

To keep things simple let's focus on California (FIPS code = 06) rather than calculating these measures for the entire US. We can use the `load_variables` function to store all 5-year ACS variables in a `tbl` and then search for population by race variables.

```{r get-tract-pop-vars, results = F}
acs19_tab_vars <- load_variables(2019, "acs5", cache = TRUE) 

# Race
acs19_tab_vars %>% 
  filter(concept == "HISPANIC OR LATINO ORIGIN BY RACE") %>% 
  View()

# Education
acs19_tab_vars %>% 
  filter(str_detect(concept, "EDUCATION"),
         str_detect(concept, "WHITE ALONE, NOT HISPANIC OR LATINO|BLACK")) %>% 
  distinct() %>% 
  View()

# Employment
acs19_tab_vars %>% 
  filter(str_detect(concept, "EMPLOYMENT"),
         str_detect(concept, "WHITE ALONE, NOT HISPANIC OR LATINO|BLACK"),
         str_detect(label, "Civilian")) %>% 
  distinct() %>% 
  View()

# Household income
acs19_tab_vars %>% 
  filter(str_detect(concept, "HOUSEHOLD INCOME"),
         str_detect(concept, "WHITE ALONE, NOT HISPANIC OR LATINO|BLACK")) %>% 
  distinct() %>% 
  View()

# Homeownership status (Census calls this tenure)
acs19_tab_vars %>% 
  filter(str_detect(concept, "TENURE"),
         str_detect(concept, "WHITE ALONE, NOT HISPANIC OR LATINO|BLACK")) %>% 
  distinct() %>% 
  View()
```

It looks like these are the relevant variables for each unidimensional measure:

-   Segregation: `B03002_003` and `B02001_003` (count of White and Black residents)

-   Education:

    -   White NH college graduation rate ages 25+ = (`C15002H_006` + `C15002H_011`) / `C15002H_001`
    -   Black college graduation rate ages 25+ = (`C15002B_006` + `C15002B_011`) / `C15002B_001`

-   Employment:

    -   White NH employment rate among civilians in labor force ages 16-64 = (`C23002H_007` + `C23002H_020`) / (`C23002H_006` + `C23002H_019`)
    -   Black employment rate among civilians in labor force ages 16-64 = (`C23002B_007` + `C23002B_020`) / (`C23002B_006` + `C23002B_019`)

-   Income:

    -   Count of White NH households with income \$100k+ = `B19001H_014` + `B19001H_015` + `B19001H_016` + `B19001H_017`
    -   Count of Black households with income under \$25k = `B19001B_002` + `B19001B_003` + `B19001B_004` + `B19001B_005`

-   [Homeownership](https://www.census.gov/quickfacts/fact/note/US/HSG445221#:~:text=The%20homeownership%20rate%20is%20computed,occupied%20housing%20units%20or%20households.&text=This%20Fact%20is%20based%20on,by%20the%20U.S.%20Census%20Bureau.):

    -   White NH homeownership rate = `B25003H_002` / `B25003H_001`

    -   Black homeownership rate = `B25003B_002` / `B25003B_001`

```{r check-vars-codebook}
c("B03002_003", "B02001_003", # population
"C15002H_006", "C15002H_011", "C15002H_001", # education 
"C15002B_006", "C15002B_011",  "C15002B_001",
"C23002H_007", "C23002H_020", "C23002H_006", "C23002H_019", # employment
"C23002H_007", "C23002H_020", "C23002H_006", "C23002H_019",
"B19001H_014", "B19001H_015", "B19001H_016", "B19001H_017", # income
"B19001B_002", "B19001B_003", "B19001B_004", "B19001B_005",
"B25003H_002", "B25003H_001", # homeownership  
"B25003B_002", "B25003B_001") %in% acs19_tab_vars$name 
```

### Calculate unidimensional measures for all PUMAs in California

For the dissimilarity index, we will need to pull data at both the tract and PUMA level. For all other measures, we can just get data at the PUMA level.

#### 1. Dissimilarity index

Our variables of interest for the dissimilarity index are `B03002_003` and `B02001_003`: Black and White population counts. We'll get these variable values for all tracts and PUMAs in California.

```{r get-tract-race-pops, message = F}
race_pop_tract <- 
  get_acs(
    geography = "tract",
    #variables = c("B03002_003", "B02001_003"),
    variables = c("B03002_003", "B03002_004", "B03002_014"),
    state = "06",
    survey = "acs5",
    year = 2019) %>% 
  rename(
    tract_name = NAME,
    tract_est = estimate,
    tract_moe = moe
  )

race_pop_puma <- 
  get_acs(
    geography = "public use microdata area",
    #variables = c("B03002_003", "B02001_003"),
    variables = c("B03002_003", "B03002_004", "B03002_014"),
    state = "06",
    survey = "acs5",
    year = 2019) %>% 
  rename(
    puma_name = NAME,
    puma_est = estimate,
    puma_moe = moe
  )
```

We also need a crosswalk between PUMAs and census tracts. The 2015-2019 ACS used the 2010 PUMA definitions - we can use the 2010 PUMA-tract crosswalk available from the Census [here](https://www2.census.gov/geo/docs/maps-data/data/rel/2010_Census_Tract_to_2010_PUMA.txt). We need to modify the TRACTCE and PUMA5CE labels to to match the GEOID variable syntax in our data (tract GEOID has 11 characters and PUMA GEOID has 7).

```{r load-xwalk}
# load in 2010 tract-PUMA cross walks
xwalk <- fread(
  "https://www2.census.gov/geo/docs/maps-data/data/rel/2010_Census_Tract_to_2010_PUMA.txt",
  colClasses = 'character',
  data.table = T) %>% 
  filter(STATEFP == "06")

# paste FIPS codes into tract and PUMA variables to match GEOID length
xwalk <- xwalk %>% 
  mutate(TRACT11 = paste0(STATEFP, COUNTYFP, TRACTCE),
         PUMA7 = paste0(STATEFP, PUMA5CE)) %>% 
  dplyr::select(TRACT11, PUMA7)
```

Now, let's get the `PUMA7` ID from `xwalk` for each tract in `race_pop_tract` by matching on `TRACT11`. We'll do this in a new `data.frame` called `joined`. We'll then pull in the corresponding `PUMA7` population estimates from `race_pop_puma`.

```{r join-puma-data}
# join in PUMA IDs from xwalk
joined <- left_join(
  race_pop_tract, 
  xwalk,
  by = c("GEOID" = "TRACT11")) %>% 
  rename("TRACT11" = "GEOID")

# join in PUMA population data from race_pop_puma
joined <- left_join(
  joined,
  race_pop_puma,
  by = c("PUMA7" = "GEOID", 
         "variable")) 
```

Note that there is one tract in `joined` (`06037137000`) that is missing its corresponding PUMA. It looks like this tract is present in the 2020 crosswalk, however (available [here](https://www2.census.gov/geo/docs/maps-data/data/rel2020/2020_Census_Tract_to_2020_PUMA.txt)), and rolls up to a 2010 PUMA. Apparently this was a mistake (more information about this [issue](https://www.census.gov/programs-surveys/acs/technical-documentation/table-and-geography-changes/2012/geography-changes.html) is available from the Census). *Note: a version of the 2010-2020 PUMA equivalency file is also available [here](https://mcdc.missouri.edu/data/corrlst/puma2010-to-puma2020.csv); we don't need it for this analysis though.*

```{r analyze-missings, message = F}
# get vector of unique tract IDs from joined that are missing corrsponding PUMA7
missings <- joined %>% 
  filter(is.na(PUMA7)) %>% 
  distinct(TRACT11) %>% 
  deframe()

# load 2020 xwalk and check to make sure that it contains the missing tract
xwalk20 <- fread(
  "https://www2.census.gov/geo/docs/maps-data/data/rel2020/2020_Census_Tract_to_2020_PUMA.txt",
  colClasses = 'character', 
  data.table = T) %>% 
  filter(STATEFP == "06") %>% 
  mutate(TRACT11 = paste0(STATEFP, COUNTYFP, TRACTCE),
         PUMA7 = paste0(STATEFP, PUMA5CE)) %>% 
  dplyr::select(TRACT11, PUMA7)

missings %in% xwalk20$TRACT11 
```

Let's fix our crosswalk so all tracts roll up into a PUMA. In correcting this issue, the Census removed tract `06037930401` (which rolls up to the same PUMA). We can simply overwrite `06037930401` with `06037137000` in our crosswalk, since they correspond to the same PUMA.

```{r fix-missings}
# append that row from xwalk20 to xwalk and re-run the joins
#xwalk <- xwalk20 %>% filter(TRACT11 %in% missings) %>% rbind(xwalk)

xwalk[xwalk == "06037930401"] <- "06037137000"

joined <- left_join(
  race_pop_tract, 
  xwalk,
  by = c("GEOID" = "TRACT11")) %>% 
  rename("TRACT11" = "GEOID")

joined <- left_join(
  joined,
  race_pop_puma,
  by = c("PUMA7" = "GEOID", 
         "variable")) 
```

Now we can calculate the PUMA-level segregation index described above, which as a reminder is: $$ D = \frac{1}{2} \sum^{n}_{i=1} \big|\frac{w_i}{W_T} - \frac{b_i}{B_T}\big|$$

Let's also rename the values of `joined$variable` from `B03002_003` and `B02001_003` to `white` and `black` in order to make it easier to keep track.

```{r calc-puma-segregation}
joined <- joined %>% 
  mutate(
    variable = case_when(
      variable == "B03002_003" ~ "white",
      variable == "B03002_004" ~ "black_nh",
      variable == "B03002_014" ~ "black_h",
    )
  )

# calculate wi/WT and bi/BT proportions for each tract
joined <- joined %>% mutate(prop = tract_est / puma_est)

# roll up variables to tract level
joined <- joined %>% 
  dplyr::select(-c("tract_est", "tract_moe", "puma_est", "puma_moe")) %>% 
  pivot_wider(
    names_from = variable,
    names_glue = "{variable}_prop",
    values_from = prop
  )

# sum black_nh and black_h proportions
joined <- joined %>%
  mutate(black_prop = black_nh_prop + black_h_prop) %>% 
  dplyr::select(-c("black_nh_prop", "black_h_prop"))

# calculate wi/WT - bi/BT for each tract
joined <- joined %>% mutate(prop_diff = abs(white_prop - black_prop))

# roll up to PUMA level and calculate dissimilarity index
joined_summ <- joined %>% 
  group_by(PUMA7) %>% 
  summarize(dissim_index = 1/2 * sum(prop_diff))

options(scipen=999) # turn off scientific notation
```

Let's map the dissimilarity index:

```{r mapping, message = F}
ca_puma_map <- pumas("CA") %>% 
  left_join(joined_summ, by = c("GEOID10" = "PUMA7"))

mapview(ca_puma_map, zcol = "dissim_index")
```

#### 2. Education inequity

Let's now move to the education inequity dimension of the MMSR. As a reminder, we need the college graduation rate of White (`(C15002H_006` + `C15002H_011`) / `C15002H_001` ) and Black ((`C15002B_006` + `C15002B_011`) / `C15002B_001`) adults over age 25. These formulas sum the counts of male and female college graduates and then divide by the total population count. We'll then take the White-to-Black ratio of college graduation rates in each PUMA to get the education dimension of the MMSR. Finally, we'll `join` these into our summary `data.frame`, `joined_summ`.

```{r calc-edu-dim}
# first pull in the relevant variables into a df called edu_puma
edu_puma <- 
  get_acs(
    geography = "public use microdata area",
    variables = c("C15002H_006", "C15002H_011", "C15002H_001",
                  "C15002B_006", "C15002B_011", "C15002B_001"),
    state = "06",
    survey = "acs5",
    year = 2019) %>% 
  rename(PUMA7 = GEOID) %>% 
  dplyr::select(-moe)

# pivot from long to wide format for calculations within each PUMA
edu_puma <- edu_puma %>% 
  pivot_wider(
    names_from = variable,
    values_from = estimate
  )

# calculate the white-black ratio in college graduation rates
edu_puma <- edu_puma %>% 
  mutate(
    edu_dim = 
      ((C15002H_006 + C15002H_011) / C15002H_001) / 
      ((C15002B_006 + C15002B_011) / C15002B_001)
    ) %>% 
  dplyr::select(PUMA7, NAME, edu_dim)

# join values into joined_summ
joined_summ <- left_join(joined_summ, edu_puma, by = "PUMA7") %>% 
  dplyr::select(PUMA7, NAME, dissim_index, edu_dim)
```

#### 3. Employment inequity

For the employment inequity dimension of the MMSR, we need the employment rate of White (`C23002H_007` + `C23002H_020`) / (`C23002H_006` + `C23002H_019`) and Black (`C23002B_007` + `C23002B_020`) / (`C23002B_006` + `C23002B_019`) civilians ages 16-64 in the labor force. These formulas sum the counts of employed male and female civilians ages 16-64 and then divide by the total count of civilians 16-64 in the labor force. Like with the education dimension, we calculate the White-to-Black rate ratio in each PUMA, and `join` these into `joined_summ`.

```{r calc-emp-dim}
# first pull in the relevant variables into a df called emp_puma
emp_puma <- 
  get_acs(
    geography = "public use microdata area",
    variables = c("C23002H_007", "C23002H_020", "C23002H_006", "C23002H_019",
                  "C23002B_007", "C23002B_020", "C23002B_006", "C23002B_019"),
    state = "06",
    survey = "acs5",
    year = 2019) %>% 
  rename(PUMA7 = GEOID) %>% 
  dplyr::select(-c("NAME", "moe"))

# pivot from long to wide format for calculations within each PUMA
emp_puma <- emp_puma %>% 
  pivot_wider(
    names_from = variable,
    values_from = estimate
  )

# calculate the white-black ratio in employment rates
emp_puma <- emp_puma %>% 
  mutate(
    emp_dim = 
      ((C23002H_007 + C23002H_020) / (C23002H_006 + C23002H_019)) / 
      ((C23002B_007 + C23002B_020) / (C23002B_006 + C23002B_019)) 
    ) %>% 
  dplyr::select(PUMA7, emp_dim)

# join values into joined_summ
joined_summ <- left_join(joined_summ, emp_puma, by = "PUMA7")
```

#### 4. Income inequity

To represent income inequity, we will use the index of concentration at the extremes (ICE). To calculate the ICE, we take the difference in number of White households with incomes of \$100k+ (`B19001H_014` + `B19001H_015` + `B19001H_016` + `B19001H_017`) and Black households with incomes under \$25k (`B19001B_002` + `B19001B_003` + `B19001B_004` + `B19001B_005`), and divide by the total number of White (`B19001H_001`) and Black (`B19001B_001`) households in the PUMA. The ICE should range from -1 to 1.

```{r calc-ice-dim}
# first pull in the relevant variables into a df called ice_puma
ice_puma <- 
  get_acs(
    geography = "public use microdata area",
    variables = c("B19001H_014", "B19001H_015", "B19001H_016", "B19001H_017", "B19001H_001",
                  "B19001B_002", "B19001B_003", "B19001B_004", "B19001B_005", "B19001B_001"),
    state = "06",
    survey = "acs5",
    year = 2019) %>% 
  rename(PUMA7 = GEOID) %>% 
  dplyr::select(-c("NAME", "moe"))

# pivot from long to wide format for calculations within each PUMA
ice_puma <- ice_puma %>% 
  pivot_wider(
    names_from = variable,
    values_from = estimate
  )

# calculate the ICE
ice_puma <- ice_puma %>% 
  mutate(
    ice_dim = 
      ((B19001H_014 + B19001H_015 + B19001H_016 + B19001H_017) - 
         (B19001B_002 + B19001B_003 + B19001B_004 + B19001B_005)) /
      (B19001H_001 + B19001B_001)
    ) %>% 
  dplyr::select(PUMA7, ice_dim)

# join values into joined_summ
joined_summ <- left_join(joined_summ, ice_puma, by = "PUMA7")
```

#### 5. Homeownership inequity

For the last dimension, homeownership inequity, we will simply calculate the ratio of White (`B25003H_002` / `B25003H_001`) to Black (`B25003B_002` / `B25003B_001`) owner-occupancy (homeownership) rates.

```{r calc-home-dim}
# first pull in the relevant variables into a df called emp_puma
home_puma <- 
  get_acs(
    geography = "public use microdata area",
    variables = c("B25003H_002", "B25003H_001", "B25003B_002", "B25003B_001"),
    state = "06",
    survey = "acs5",
    year = 2019) %>% 
  rename(PUMA7 = GEOID) %>% 
  dplyr::select(-c("NAME", "moe"))

# pivot from long to wide format for calculations within each PUMA
home_puma <- home_puma %>% 
  pivot_wider(
    names_from = variable,
    values_from = estimate
  )

# calculate the white-black ratio in employment rates
home_puma <- home_puma %>% 
  mutate(
    home_dim = 
      (B25003H_002 / B25003H_001) / (B25003B_002 / B25003B_001)
    ) %>% 
  dplyr::select(PUMA7, home_dim)

# join values into joined_summ
joined_summ <- left_join(joined_summ, home_puma, by = "PUMA7")
```

### Calculate MMSR for all PUMAs in California

Now that we have the unidimensional measures for each of the five dimensions for all PUMAs in California, we'll move on to the construction of the multidimensional measure. The bulk of this involves conducting a [latent class analysis](https://www.theanalysisfactor.com/what-is-latent-class-analysis/) (LCA), which allows us to identify subgroups (latent classes) of PUMAs that share like characteristics, namely, similar values of the unidimensional measures we've just calculated.

Before we move into the LCA, we'll need to dichotomize each measure into high versus low based on the sample median for each dimension. Let's set values that are greater than the median to 1 (to represent high), and 0 otherwise (to represent low).

```{r dichot-dims}
joined_summ <- joined_summ %>% 
  mutate(
    dissim_bin = case_when(
      dissim_index > median(dissim_index) ~ 1,
      TRUE ~ 0), 
    edu_bin = case_when(
      edu_dim > median(edu_dim) ~ 1,
      TRUE ~ 0), 
    emp_bin = case_when(
      emp_dim > median(emp_dim) ~ 1,
      TRUE ~ 0), 
    ice_bin = case_when(
      ice_dim > median(ice_dim) ~ 1,
      TRUE ~ 0), 
    home_bin = case_when(
      home_dim > median(home_dim) ~ 1,
      TRUE ~ 0)
    )
```

```{r}
joined_summ %>% 
  dplyr::filter(dissim_bin == 1 & 
                  edu_bin == 1 & 
                  emp_bin == 1 & 
                  ice_bin == 1 & 
                  home_bin == 1) %>% 
  pull(NAME)

joined_summ %>% 
  dplyr::filter(dissim_bin == 0 & 
                  edu_bin == 0 & 
                  emp_bin == 0 & 
                  ice_bin == 0 & 
                  home_bin == 0) %>% 
  pull(NAME)
```

```{r}
joined_summ <- joined_summ %>% 
  mutate(dim_sums = 
           dissim_index / max(dissim_index) + 
           edu_dim / max(edu_dim)  + 
           emp_dim / max(emp_dim)  + 
           ice_dim / max(ice_dim)  + 
           home_dim / max(home_dim) ,
         bin_sums = dissim_bin + edu_bin + emp_bin + ice_bin + home_bin)

ca_puma_map <- pumas("CA") %>% 
  left_join(joined_summ, by = c("GEOID10" = "PUMA7"))

mapview(ca_puma_map, zcol = "dim_sums")
```

#### Run latent class analysis
